---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BRGenomics

<!-- badges: start (versions currently manual...) -->

[![platforms](https://img.shields.io/badge/platforms-linux%20%7C%20osx%20%7C%20win-yellow.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/BRGenomics/)
[![release](https://img.shields.io/badge/release%20version-1.0.3-navy.svg)](https://www.bioconductor.org/packages/BRGenomics)
[![devel](https://img.shields.io/badge/devel%20version-1.1.3-orange.svg)](https://github.com/mdeber/BRGenomics)
[![updated](http://www.bioconductor.org/shields/lastcommit/devel/bioc/BRGenomics.svg)](http://bioconductor.org/checkResults/devel/bioc-LATEST/BRGenomics/)
[![last commit](https://img.shields.io/github/last-commit/mdeber/BRGenomics.svg)](https://github.com/mdeber/BRGenomics/commits/master)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/BRGenomics.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/BRGenomics/)

<!-- badges: end -->

Efficient tools for the analysis of high-resolution genomics data in R.

Explore the documentation: https://mdeber.github.io/

# Installation

As of Bioconductor 3.11 (release date April 28, 2020), BRGenomics can be 
installed directly from Bioconductor:

```{r, eval = FALSE}
# install.packages("BiocManager")
BiocManager::install("BRGenomics")
```

Alternatively, the latest development version can be installed from [GitHub](
https://github.com/mdeber/BRGenomics):

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("mdeber/BRGenomics@R3")
```

_BRGenomics (and Bioconductor 3.11) require R version 4.0 (release April 24, 
2020). Installing the `R3` branch, as shown above, can be installed under
R >= 3.5_

If you install the development version from Github and you're using Windows, 
[Rtools for Windows](https://cran.rstudio.com/bin/windows/Rtools/) is required.

## Features

See the [documentation website](https://mdeber.github.io/), which includes an 
introductory vignette, as well as the documentation for currently implemented 
functions, complete with demonstrative example code. The package currently 
includes example PRO-seq data^[1]^.

## Limitations for Windows users

* No support for parallel/multicore processing
* No support for import bigWig files

## Possible future changes

* Convert several functions into S4 generics
    + _Note that in future versions, the names of the first argument of
    certain functions may change, but their order/utility will not. I.e., 
    functions with with the first argument `dataset.gr` may become `x`_
* Formalize support for `GRangesList` objects (should be already complete)
* Write methods for `BigWigFile`/`BigWigFileList` objects (to avoid loading 
data into memory)
* (Possibly) use `GPos` objects by default
* Vectorize `genebodies()` function (like many `GenomicRanges` functions are)

---

1. Hojoong Kwak, Nicholas J. Fuda, Leighton J. Core, John T. Lis (2013). 
Precise Maps of RNA Polymerase Reveal How Promoters Direct Initiation and 
Pausing. *Science* __339__(6122): 950–953. 
https://doi.org/10.1126/science.1229386
