---
title: "Basepair-Resolution Analysis with BRGenomics"
subtitle: "_Straightforward and efficient tools for analyzing genomics data_ [EARLY DRAFT]"
author:
  name: Mike DeBerardine
  email: mike.deberardine@gmail.com
package: BRGenomics
output:
  BiocStyle::html_document
abstract: |
  BRGenomics is designed to help users avoid code repetition by providing efficient and tested functions to accomplish common, discrete tasks in the analysis of high-throughput sequencing data. The included functions are geared toward analyzing basepair-resolution sequencing data, the properties of which are exploited to increase performance and user-friendliness. We leverage standard Bioconductor methods and classes to maximize compatability with its rich ecoystem of bioinformatics tools, and we aim to make BRGenomics sufficient for most post-alignment data processing. Common data processing and analytical steps are turned into fast-running one-liners that can be simultaneously applied across numerous datasets. BRGenomics is fully-documented, and we aim it to be beginner-friendly.
  
vignette: |
  %\VignetteIndexEntry{Overview of BRGenomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Motivation

This package is designed to:

* Replace the use of command-line utilities for most post-alignment processing, e.g. `bedtools` and `deeptools`
* Be easy-to-use and easy-to-install, without requiring external dependencies, e.g. `hitslib` or the kent source utilities from the UCSC genome browser
* Allow users to string together common analysis pipelines with simple, fast-running one-liners
* Avoid code repetition by providing tested and validated code
* Exploit the properties of basepair-resolution data to optimize performance and increase user-friendliness
* Use process forking to make use of multicore processors 
* Maximize compatability with Bioconductor's rich ecosystem of analysis software, in addition to leveraging the traditional strengths of R in statistics and data visualization
* Fully replace the `bigWig` R package

# Features

* Process and import bedGraph, bigWig, and bam files quickly and easily, with several pre-configured defaults for typical uses
* Count and filter spike-in reads
* Calculate spike-in normalization factors using several methods and options, including options for batch normalization
* Count reads by regions of interest
* Count reads at positions within regions of interest, at single-base resolution or in larger bins, and generate count matrices for heatmapping
* Calculate bootstrapped signal (e.g. readcount) profiles with confidence intervals (i.e. meta-profiles)
* Modify gene regions (e.g. extract promoters or genebody regions) using a single simple and straightforward function
* Conveniently and efficiently call `DESeq2` to calculate differential expression in a manner that is robust to global changes^[Avoid the default behavior of calculating genewise dispersion across all samples present, which is invalid if any experimental condition causes broad changes]
  + Use non-contiguous genes in `DESeq2` analysis, e.g. to exclude of specific sites/peaks from the analysis (not usually supported by DESeq2)
  + Efficiently generate results across a list of comparisons
* Support for blacklisting throughout, and proper accounting of blacklisted sites in relevant calculations
* Users interact with an intuitive and computationally efficient data structure (the "basepair resolution `GRanges`" object), which is already supported by a rich, user-friendly suite of tools that greatly simplify working with datasets and annotations

### Coming Soon

Data processing:

* Summarizing and plotting replicate correlations
* Function to use random read sampling to assess if sequencing depth sufficient to stabilize arbitrary calculations (so a user can supply anonymous function to calculate things like rank expression, power analysis or differential expression by DESeq2, pausing indices, etc.)

Signal counting and analysis:

* Two-stranded meta-profile calculations
* Automated generation of a list of DESeq2 comparisons using all possible combinations; all possible permutations; or by defining a simple hierarchy of each-vs-one comparisons

# Installation

Install development version from [GitHub](https://github.com/mdeber/BRGenomics):

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mdeber/BRGenomics")
```

If you're using Windows, [Rtools for Windows](https://cran.rstudio.com/bin/windows/Rtools/) is required.

# Getting started

The package ships with an example dataset of PRO-seq data from Drosophila melanogaster^[Hojoong Kwak, Nicholas J. Fuda, Leighton J. Core, John T. Lis (2013). Precise Maps of RNA Polymerase Reveal How Promoters Direct Initiation and Pausing. _Science_  __339__(6122): 950–953. https://doi.org/10.1126/science.1229386]. PRO-seq is a basepair-resolution method that uses 3'-end sequencing of nascent RNA to map the locations of actively engaged RNA polymerases. To keep the dataset small, we've only included reads mapping to the fourth chromosome^[Chromosome 4 in Drosophila, often referred to as the "dot" chromosome, is very small and contains very few genes]. The data is included in several formats, all of which can be derived from the included bam file. 

# Working With Bam Files

The `import_bam` function provides a number of options for filtering and processing bam files. Because PRO-seq data is sequenced in the 3'-to-5' direction of the original RNA molecule, we'll use the `revcomp` option to reverse-complement all the input reads. We'll also set a minimum MAPQ score of 20:

```{r, message=FALSE}
library(BRGenomics)

# path of included bam file
bfile <- system.file("extdata", "PROseq_dm6_chr4.bam", 
                     package = "BRGenomics")

ps_reads <- import_bam(bfile, mapq = 20, revcomp = TRUE, paired_end = FALSE)
ps_reads
```

Files are imported as `GRanges` objects from the `r Biocpkg("GenomicRanges")` package. GRanges objects are very easy to work with, and are supported by a plethora of useful functions and packages.

---

_To learn more about GRanges objects, including a general overview of their components, see the useful vignette [_An Introduction to the GenomicRanges Package_](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.html). Alternatively, see the archived materials from the 2018 Bioconductor workshop [_Solving Common Bioinformatic Challenges Using GenomicRanges_](https://bioconductor.github.io/BiocWorkshops/solving-common-bioinformatic-challenges-using-genomicranges.html). Note that this package will implement and streamline a number of common operations, but users should still have a basic familiarity with GRanges objects._

---

By default, `import_bam` combines identical reads into the same range, and uses the `score` metadata column to represent how many times that read occurs. This means that the total number of reads in the dataset is equal to the total score:

```{r, collapse = TRUE}
sum(score(ps_reads))
```

Alternatively, you can import each read as its own range by setting `field = NULL`:

```{r}
reads_expanded <- import_bam(bfile, mapq = 20, revcomp = TRUE, 
                             field = NULL, paired_end = FALSE)
ps_reads[1:8]
reads_expanded[1:8]
```

```{r}
length(reads_expanded) == sum(score(ps_reads))
```

# Basepair-Resolution GRanges Objects

The base immediately before the 3' end of a mapped PRO-seq read gives the position of an RNA polymerase at basepair-resolution. The `import_bam` function also includes options for selecting specific bases of each read, which are applied after reverse-complementing. Here, we reverse complement, take the 3' end of the read, and then shift that base upstream by 1 base:

```{r}
ps <- import_bam(bfile, 
                 mapq = 20, 
                 revcomp = TRUE,
                 shift = -1,
                 trim.to = "3p",
                 paired_end = FALSE)
ps
```

You may have noticed that this GRanges object isn't the same length as when we imported the entire reads. That's because after trimming, identical positions were collapsed together, and the `score` column contains the readcounts at each position.

The returned data is a GRanges object in which all ranges represent a single base, and in which a metadata field (usually `score`) contains readcounts. This data structure underpins much of this package, and we refer to it as a __basepair-resolution GRanges object__^[The `GPos` class is designed for a similar purpose, but we currently do not use it. There is a note on this in the documentation for `makeGRangesBRG`].

---

_bigWig files, as well as `coverage` objects in `GenomicRanges`, use a form of compression in which adjacent positions sharing the same signal are combined. For whole-read coverage methods, this run-length encoding enhances performance and efficiency. However, basepair resolution data is not nearly as smooth, and its rare that adjacent bases share the same signal. With basepair resolution data, combining adjacent positions increases complexity, while providing a negligible storage benefit._

---

Splitting up each base into its own range (its own index) results in a number of downstream computational benefits. Perhaps more importantly, however, is the added simplicity for the user: each index of a basepair-resolution GRanges object addresses a single genomic position and its associated signal. 

```{r, collapse=TRUE}
length(ps) # number of unique positions in dataset
sum(score(ps)) # total number of reads
sum(score(ps) == 1) # number of sites with only a single read
```

```{r}
subset(ps, score > 10) # sites with more than 10 reads
```

We can write a simple test to determine if a GRanges object is "basepair-resolution":

```{r, collapse=TRUE}
all(width(ps) == 1)
```

# Example: BAM to bigWig Conversion

The `import_bam` function provides several options for processing aligned reads, and in conjunction with functions from the `r Biocpkg("rtracklayer")` package, we can easily write a post-alignment pipeline for generating bigWig files for viewing in a genome browser. Here, we use the `import_bam_PROseq` function, which has preset defaults for PRO-seq data. For other pre-configured importation functions, see the documentation for the `import_bam` function.

```{r, eval=FALSE}
ps <- import_bam_PROseq(bfile, mapq = 20, paired_end = FALSE)

ps_plus <- subset(ps, strand == "+")
ps_minus <- subset(ps, strand == "-")
score(ps_minus) <- -score(ps_minus) # make minus-strand scores negative

rtracklayer::export.bw(ps_plus, "~/Data/PROseq_plus.bw")
rtracklayer::export.bw(ps_minus, "~/Data/PROseq_minus.bw")
```

# Spike-in Normalizing Data

# Working With Annotations

<!--
* Importing genelists
* Modifying genelists using genebodies for instance
* Finding max positions
* (Use example, find max position, then make heatmap centered on it)
-->

# Heatmaps

<!-- (Use image, and use ComplexHeatmap; and show reshape2::melt to use ggplot or lattice) -->

# Metaplots

<!-- (Use R plotting, and ggplot) -->

# Efficient Analysis of Multiple Datasets

To efficiently work with several datasets, we recommend storing the GRanges objects within a standard, named list, e.g. `grl <- list(a_rep1 = gr1, b_rep1 = gr2, ...)`. A named list like this can be passed as an argument to nearly every function in BRGenomics, and many functions will automatically return dataframes, or melted dataframes that use the list names as the sample names.

BRGenomics does not currently support the use of `GRangesList` or `CompressedGRangesList` classes for grouping multiple datasets.^[The `GRangesList` and `CompressedGRangesList` classes, particularly the latter, often cause performance penalties. But more importantly, these classes are often used to store fundamentally "single" objects that have constituent subgroups, like exons organized by transcript. Especially given their wide support by Bioconductor packages, using them as a simple container for multiple GRanges objects creates the potential for undesired method dispatch, and moreover reduces their compatability with non-Bioconductor list-focused methods, like `mclapply`.]

<!--

(Recommend storing GRanges objects within standard R list elements. 

(And there is a GRangesList class and a compressed version, but we've found 
substantial performance penalties in using them. Note that these classes may scale
better for analyzing a large number of datasets simultaneously, when memory is
limiting. They're currently not supported).

## Multiplexed GRanges

(Depending on the dataset, it may make sense to combine many datasets into a single
GRanges object, with readcounts for each dataset contained in the `mcols` metadata.
We've termed these "multiplexed GRanges" objects, and they are supported throughout
the package.)

((make one and demonstrate))

(While this format is tempting, and on first inspection would appear to be more
efficient, in our experience there is a substantial performance penalty when
used with sparse data. With sparse data, like PRO-seq, it's more efficient to have
a number of smaller GRanges objects than to have a single very large one containing
numerous metadata columns containing mostly zeroes. This may change if Bioconductor
core classes implement future support for efficient handling of sparse matrices.)

-->

# Saving Binary R Files

While the handling of GRanges data in BRGenomics is relatively fast, the initial importation of bigWig or bedGraph files as GRanges objects remains a noticeable bottleneck. This may be tolerable for interactive workflows, in which data is imported once before undergoing lengthy analysis, but the bottleneck is a significant detriment for users who regularly import data.

To avoid this bottleneck, users can save reusable data structures as binary R files, which effectively save the memory state of R objects. Not only are these objects rapidly reloaded into memory upon importation, but they have the added benefit of saving the user from repeating data formatting (like naming).

<!--(maybe footnote the save/read vs. saveRDS/readRDS) -->

Any R object can be saved to storage using the `save` or `saveRDS` commands, and re-imported using the `load` or `readRDS` commands. The `saveRDS`/`readRDS` commands work like other data exporting and importing commands, while the `read`/`load` commands maintain the original name of the objects. For instance, if you `save(PROseq, file = "~/ps.RData")`, and in a new R session run `read("~/ps.RData")`, a new object called `PROseq` will be created in your new environment.^[Note that this is the same way that RStudio saves your current working environment to disk, i.e. it saves the entire environment into an RData file.]


# DESeq

DESeq2's default treatment of data relies on the assumption that genewise estimates of dispersion are largely unchanged across samples. While this assumption holds for a typical RNA-seq data, it can be violated if there exist _any_ conditions represented in the `DESeqDataSet` object that caused meaningful signal changes across a majority of regions of interest. The BRGenomics functions `getDESeqDataSet` and `getDESeqResults` are simple and flexible wrappers for making pairwise comparisons between individual samples, without relying on the assumption of globally-similar dispersion estimates.

For making simple, pairwise comparisons between individual (replicated) samples, the `getDESeqResults` follows the logic that the presence of a dataset $X$ within the `DESeqDataSet` object will not affect the comparison of datasets $Y$ and $Z$. However, while this intuition is appealing, users should note that if the globally-similar dispersions assumption _does_ hold, then DESeq2's default behavior should be more sensitive in its estimates of genewise dispersion. In this case, users can still take advantage of the convenience of the BRGenomics function `getDESeqDataSet` before calling `DESeq2::DESeq` and `DESeq2::results`.

If the globally-similar dispersions assumption is violated, but something beyond simple pairwise comparisons is desired (e.g. group comparisons or additional model terms), we note that DESeq2 can be used without "blind dispersion estimation" (see the DESeq2 manual).


<!-- (Using list) -->

# Combining Replicates

Biological replicates are best used to independently reproduce and measure 
effects, and therefore we often want to handle them separately. However, there are 
times when combining replicates can allow for more sensitive measurements, assuming
that the replicates are indistinguishable.

The `mergeGRangesData` function can be used to combine basepair-resolution GRanges objects, e.g. `a_combined <- mergeGRangesData(a_rep1, a_rep2)`

<!-- 

((Later on))

(There are other times when it makes sense to combine a large number of datasets.
This can be accomplished by passing an entire list of GRanges objects)

-->

# "Browser Shots"

# Sequence Extraction

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
devtools::session_info()
```


<!-- SCRAP -->

<!-- The `import_bam` function provides several options for processing aligned reads, and in conjunction with functions from the `r Biocpkg("rtracklayer")` package, we can easily write a post-alignment pipeline for generating bigWig files for viewing in a genome browser. -->

<!-- As an exmaple, PRO-seq data is sequenced in the antisense orientation to the original RNA molecule, the 3' end of which provides the location of the active site of an engaged RNA polymerase. Therefore, PRO-seq reads need to be reverse-complemented, and the 3' ends need to be extracted. -->
